---
title: "Parallel computations in R"
author: "Qin Li"
date: '2017.02'
output:
  html_document: default
---

#### Why am I doing this

The common motivation behind parallel computing is that something is taking too long time. When the task is repeating the same function (or bootstrapping/fitting multiple regression models), you should start executing parallel programs!

#### What are the steps

- **run R scripts on cluster**
    - basic steps:
        1. connect to the cluster with your ZCU account
            - command: 
                - `ssh your_account@zoology.ubc.ca`
                - `ssh cluster` # 6 jobs only on the head node
                - `ssh crunch**` # use another less busy node (up to 16 jobs)
            - ZCU wiki:
                - cluster access: https://zcu.zoology.ubc.ca:442/wiki/index.php/Cluster_Access
                - cluster status: https://clweb.zoology.ubc.ca/orwell/
        2. start a screen session
            - command: `screen -S your_screen_name` 
            - because you want to keep the script running even if your connection drops
            - you could create different screens with various names for multiple programming
            - then you can "shutdown your laptop, go home, log in to the cluster from a different computer, attach to the screen session, and continue to interact with your program."
        3. run your R script
            - command: `Rscript your_R_script.R`
            - make sure you've already uploaded your data and scripts *(tested)* to a certain directory
            - you can use command lines to do this (i.e., `scp`), but it will be easier and faster using a FTP client, e.g. *FileZilla*, to upload or download files
        4. detach from a screen session
            - keyboard command: `ctrl-a d`
                - press `control` and then `a`, hold them, then hit `d`
            - after this, you will return to the previous shell window and you can disconnect your client with your job running in the background
        5. attach to a running  screen
            - command to check all screens: `screen -ls`
            - command to attch one: `screen -r your_screen_name`
        6. terminate a screen
            - command: `exit`
        7. log out
            - keyboard command: `ctrl-d`
    
    - command lines in *Terminal*: an example
    
    ```{r eval=FALSE}
        $ ssh qinli@zoology.ubc.ca   # connect to the main Zoology server
        Password: # type your password
        qinli% ssh cluster    # ssh to the head-node
        ~ $ ls
        ~ $ cd the_directory_you_want_to_work
        ~/work_directory $ screen -S mywork    # 'mywork' is the name of the screen
        ~/work_directory $ Rscript xxx.R    # run R script
        
        ctrl-a d    # keyboard command to detach the screen
        
        ~/work_directory $ screen -ls    # check running/detached screens
        ~/work_directory $ screen -r mywork    # attach the screen
        ~/work_directory $ exit    # determinate the screen when the job is done
        
        ctrl-d    # keyboard command (connection to cluster closed)
        ctrl-d    # keyboard command (connection to zoology.ubc.ca closed)
    ```
        
- **parallel computing in R**
    - the `parallel` package
        - is basically about repeat the same function in parallel -- *SAVE TIME!*
        - use multiple cores:
            - calculate the number of cores and initiate cluster
                - `no_cores <- detectCores() - 1`
                - `cl <- makeCluster(no_cores)`
            - a good number of clusters is *the numbers of available cores – 1*;
                - using all cores might prevent you from doing anything else;
        
    - `parLapply`: the parallel version of `lapply`
        - learning `lapply` is the key: it takes one parameter (a vector/list), apply `FUN` to each element, and returns a list of the same length
            - use `parSapply` if you only want to return a simple value and directly get it processed as a vector/matrix
        - create your own function *involving repetition*
            - which could solve your problem that is amenable to parallelism
            - parallel version: `parLapply(cl, parameter, myFUN)`
        - variable scope
            - you need to specify what variables and libraries that you need for the parallel function to work
                - `clusterEvalQ(cl, library(xxx))`
                - `clusterExport(cl, list("variable1", "variable2"))`
            - or you have the option on Mac/Linux of using `makeCluster(no_core, type="FORK")`
                -  which automatically contains all environment variables and no need to worry about variable corruption
            
    - save results and shutdown cluster
        - `stopCluster(cl)`
        - once we are done we need to close the cluster so that resources such as memory are returned to the operating system.
    
    - a simple example: simulate 100 datasets from normal distribution
        - you can easily do this with a `for` loop or `sapply`, but here is a parallel version
        
    ```{r eval=FALSE}
    # Start up a parallel cluster
    library(parallel)
    
    # create a function to generate 1000 numbers from a normal distribution 
    # with mean = x and sd = 1
    myFun <- function(x){
        dt <- rnorm(1000,x,1)
        return(dt)
    }
    
    v.mean <- c(1:100) # 100 values for means of normal distributions
    
    # initiate cluster
    no_cores <- detectCores() - 1
    cl <- makeCluster(no_core, type="FORK")
    
    # or you need to specify variables that you need for the parallel function to work
    # clusterExport(cl, list("v.mean", "myFun"))
    
    mydata <- parLapply(cl, v.mean, myFun)
    
    # tryCatch(
    # mydata <- parLapply(cl, v.mean, myFun),
    # error = function(e) print(e)
    # ) # error handling function
    
    saveRDS(mydata, file = "mydata.rds")

    stopCluster(cl) # shutdown cluster
    ```

#### more info

- screen commands: https://www.maths.ox.ac.uk/help/faqs/programming/jobs-using-screen
- parallel computing in R
    - https://www.r-bloggers.com/how-to-go-parallel-in-r-basics-tips/
    - http://www.glennklockwood.com/data-intensive/r/lapply-parallelism.html
    - http://www.win-vector.com/blog/2016/01/parallel-computing-in-r/
- google
